---
title: "Diversion Data Request"
author: "Gwen Beebe"
date: "12/22/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}

rm(list = ls())
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(pander)
library(dplyr)
library(janitor)
library(tidyr)
library(readxl)

load("C:/Users/GwenBeebe/Documents/R/CHIP_HMIS/images/all_data.RData")
load("C:/Users/GwenBeebe/Documents/R/CHIP_HMIS/images/return_flags.RData")

ReportStart <- ymd("20191001")
ReportEnd <- ymd("20201001")

##  set program types to include in calculations
included_program_types <- c("Transitional housing", "PH - Permanent Supportive Housing (disability required for entry)",
                       "PH – Housing Only", "PH – Housing with Services (no disability required for entry)",
                       "Emergency shelter", "Homelessness Prevention", "Street outreach", "PH - Rapid Re-Housing",
                       "Safe Haven", "Coordinated Entry", "Diversion")

##  add diversion category to all_data and apply program type filter
all_data <- all_data %>%
  mutate(ProgramType = 
             if_else(ProgramName == "Diversion" 
                     | ProgramName == "IHN - Diversion"
                     | ProgramName == "YHDP - Diversion",
                     "Diversion", ProgramType)) %>%
  filter(ProgramType %in% included_program_types)

##  set time period for reporting
active_in_period <- all_data %>%
  filter(EnrollDate < ReportEnd &
           (is.na(ExitDate) | ExitDate >= ReportStart))

```

## Diversion projects in HMIS (HIP, Family Promise, HVAF, other SSVF Providers)
What information is available about people that have been assisted?
(Assumption is each project is standalone due to the differences in implementation?)

What outcomes are you seeing? (demographics)
- # of household attempted to be diverted
- # of households diverted from shelter
- # of households to shelter
- # of successfully diverted households returning to homelessness

Is there any other information or outcomes being tracked such as length of follow-up services, enrollments in other programs, whether financial assistance is being provided and how much?

Note: these percentages are by row.
```{r Diversion}

##  filter to only show most recent diversion entry by heads of household
diversion_households_active_in_period <- active_in_period %>%
  filter(ProgramType == "Diversion"
         & `Head of Household?` == "Yes")  %>%
  group_by(ClientID) %>%
  arrange(EnrollDate) %>%
  slice(1L) %>%
  ungroup()

##  add column for whether client entered shelter after entering diversion
diversion_households_enhanced <- diversion_households_active_in_period %>%
  left_join(all_data %>%
              filter(ProgramType == "Emergency shelter") %>%
              select("ClientID", "EnrollDate"),
            by = "ClientID") %>%
  group_by(ClientID) %>%
  mutate(entered_shelter = if_else(EnrollDate.x <= EnrollDate.y, 1, 0),
         entered_shelter = if_else(is.na(max(entered_shelter)), 0, max(entered_shelter)),
         entered_shelter = if_else(entered_shelter == 1, "Yes", "No")) %>%
  select(-EnrollDate.y) %>%
  arrange(EnrollDate.x) %>%
  slice(1L) %>%
  ungroup()

## display whether households accessed shelter after diversion entry by race
knitr::kable(
  diversion_households_enhanced %>%
    tabyl(RaceDesc, entered_shelter) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Race", "Entered Shelter?"),
  caption = "Households Accessing Shelter After Diversion Entry"
  )

## display whether households accessed shelter after diversion entry by ethnicity
knitr::kable(
  diversion_households_enhanced %>%
    tabyl(HUDEthnicity, entered_shelter) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Ethnicity", "Entered Shelter?"),
  caption = "Households Accessing Shelter After Diversion Entry"
  )

##  flag diversion returns by head of household
returning_from_diversion <- diversion_households_active_in_period %>%
  left_join(return_flags, by = "EnrollID") %>%
  mutate(return_flag = case_when(
    is.na(return_flag) ~ "No Permanent Exit",
          return_flag == 0 ~ "No Return Yet",
          return_flag == 1 ~ "Returned")
  )

##  display return rate for diversion by head of household
knitr::kable(
  returning_from_diversion %>%
    tabyl(return_flag) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "Households", "Percent")),
  caption = "Households Returning After Successful Diversion"
  )

```

##HMIS Data at entry and at exit
####(all people with an open entry at beginning of reporting period or an entry within the reporting period)
####(count and % where possible)

Total Households Served
- Individuals and Adults with Children households count and % 

For this section, I counted each person once. This means if they appeared in a family at all, I put them in the family category; those shown in the adults only and minor categories did not appear in any other household type (if a minor aged in and had another entry as an adult, they are shown in the adults only category). Categorization is based on age at entry.

All following sections include information for the following program types unless otherwise noted: Transitional housing, PH - Permanent Supportive Housing (disability required for entry), PH – Housing Only, PH – Housing with Services (no disability required for entry), Emergency shelter, Homelessness Prevention, Street outreach, PH - Rapid Re-Housing, Safe Haven, Coordinated Entry, Diversion.

```{r households}

##  create household labels and keep one per person
household_information <- active_in_period %>%
  group_by(CaseID) %>%
  mutate(household_type = case_when(
    min(`Age at Entry`) >= 18 ~ "Adults Only",
    max(`Age at Entry`) < 18 ~ "Minor",
    min(`Age at Entry`) < 18 &
      max(`Age at Entry`) >= 18 ~ "Family",
    TRUE ~ "Unknown"
  ),
  household_type = factor(
      household_type,
      levels = c(
        "Family",
        "Adults Only",
        "Minor",
        "Unknown"
      )
    ),
  `Head of Household?` = factor(
      `Head of Household?`,
      levels = c(
        "Yes",
        "No"
      )
    )) %>%
  ungroup() %>%
  group_by(ClientID) %>%
  arrange(household_type, `Head of Household?`) %>%
  slice(1L) %>%
  ungroup()
  
## show distinct people count by household type
knitr::kable(
  household_information %>%
    tabyl(household_type) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People by Household Type"
  )

## show distinct household count by household type
knitr::kable(
  household_information %>%
    filter(`Head of Household?` == "Yes") %>%
    tabyl(household_type) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "Households", "Percent")),
  caption = "Count of Households by Household Type"
  )

```

Total People Served
- Race
- Ethnicity 
- Age groups (APR categories)
- Disability

Race and ethnicity use the most recent reported value for each person. If a client fell into two age groups due to a birthday, the age at the time of the more recent entry was used. For disabling condition, all reported values are included.

```{r total served}

##  get disability information
disabling_conditions <- 
  read_excel("C:/Users/GwenBeebe/Documents/R/CHIP_HMIS/Data Requests/data/Disabling Conditions 12.23.20.xlsx") 

##  create disabling condition table
disability_information <- active_in_period %>%
  select(ClientID) %>%
  distinct() %>%
  left_join(disabling_conditions, by = "ClientID") %>%
  mutate_all(~ replace_na(.x, 0)) %>% 
  select(-DisablingCondition, -`History of Foster Care`, -`Felony Conviction`) %>%
  pivot_longer(!ClientID, names_to = "Disability", values_to = "status") %>%
  mutate(Disability = if_else(Disability == "Other", "Other Disability", Disability))

##  create client labels and keep one per person
client_information <- active_in_period %>%
  group_by(ClientID) %>%
  arrange(desc(`Age at Entry`)) %>%
  slice(1L) %>%
  ungroup() %>%
  left_join(disability_information %>%
              group_by(ClientID) %>%
              select(-Disability) %>%
              arrange(desc(status)) %>%
              slice(1L) %>%
              ungroup(), 
            by = "ClientID") %>%
  mutate(age_group = case_when(
    `Age at Entry` < 5 ~ "Under 5",
    `Age at Entry` <= 12 ~ "5-12",
    `Age at Entry` <= 17 ~ "13-17",
    `Age at Entry` <= 24 ~ "18-24",
    `Age at Entry` <= 34 ~ "25-34",
    `Age at Entry` <= 44 ~ "35-44",
    `Age at Entry` <= 54 ~ "45-54",
    `Age at Entry` <= 61 ~ "55-61",
    `Age at Entry` <= 105 ~ "62+",
    TRUE ~ "Unknown",
  ),
  age_group = factor(
      age_group,
      levels = c(
        "Under 5", "5-12", "13-17", "18-24",
        "25-34", "35-44", "45-54", "55-61",
        "62+", "Unknown"
      )
    ),
  DisablingCondition = case_when(
    DisablingCondition == 1 | status == 1 ~ "Disabled",
    TRUE ~ "Not Known Disabled"
  ))

##  show count of people by race
knitr::kable(
  client_information %>%
    tabyl(RaceDesc, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People by Race"
  )

##  show count of people by ethnicity
knitr::kable(
  client_information %>%
    tabyl(HUDEthnicity, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People by Ethnicity"
  )

##  show count of people by age group
knitr::kable(
  client_information %>%
    tabyl(age_group, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People by Age Group"
  )

##  show count of people by disability status
knitr::kable(
  client_information %>%
    tabyl(DisablingCondition, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People by Disability Status"
  )

knitr::kable(
  disability_information %>%
  group_by(Disability) %>%
  summarise(People = sum(status)) %>%
  mutate(Percent = paste0(round(People/nrow(client_information) * 100, 1), "%")),
  caption = "Count of People by Disability Type"
  )


```


##Race/Ethnicity breakdown 
- Race by age groups and ethnicity by age groups (wish list for early 2021) 
- Race/age group and disability or no disability (UDE -  (Y/N))

Note: these percentages are by table.
```{r race and ethnicity}

##  show count of people by race and age group
knitr::kable(
  client_information %>%
    tabyl(RaceDesc, age_group, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("all") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    adorn_title("top", "Race", "Age Group"),
  caption = "Count of People by Race and Age"
  )

##  show count of people by ethnicity and age group
knitr::kable(
  client_information %>%
    tabyl(HUDEthnicity, age_group, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("all") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    adorn_title("top", "Ethnicity", "Age Group"),
  caption = "Count of People by Ethnicity and Age"
  )

##  show count of people by race and disability status
knitr::kable(
  client_information %>%
    tabyl(RaceDesc, DisablingCondition, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("all") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    adorn_title("top", "Race", ""),
  caption = "Count of People by Race and Disability Status"
  )

##  show count of people by ethnicity and disability status
knitr::kable(
  client_information %>%
    tabyl(HUDEthnicity, DisablingCondition, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("all") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    adorn_title("top", "Ethnicity", ""),
  caption = "Count of People by Ethnicity and Disability Status"
  )

##  show count of people by age group and disability status
knitr::kable(
  client_information %>%
    tabyl(age_group, DisablingCondition, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("all") %>%
    adorn_pct_formatting() %>%
    adorn_ns() %>%
    adorn_title("top", "Age Group", ""),
  caption = "Count of People by Age and Disability Status"
  )

```

Distribution at entry prior living situation and distribution at exit destination (All)
Within this dataset
- Destinations of households with a prior living situation of family and friends (perm and temporary)
- Prior living situation Jail (3.917, field 7) by race and ethnicity

Distribution of all prior residences and race/ethnicty across those entering from incarceration only includes entries within reporting period. The high rate of data not collected is due to shelter, and the high rate of nulls is due to coordinated entry and shelter. These numbers are per entry and not per client, because one client could have multiple destinations and prior residences within any given reporting period.

```{r residences}

## filter to only include entries on or after the report start date
entries_during_period <- active_in_period %>%
  filter(EnrollDate >= ReportStart) %>%
  mutate(`Prior Living Situation(43)` = 
           if_else(!is.na(`Prior Living Situation(43)`), `Prior Living Situation(43)`, "NA"))

##  show distribution of prior residences across entries
knitr::kable(
  entries_during_period %>%
    tabyl(`Prior Living Situation(43)`) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "Entries", "Percent")),
  caption = "Count of Entries by Prior Residence"
  )

##  show distribution of exit destinations across entries from couching situations
knitr::kable(
  active_in_period %>%
    filter((`Prior Living Situation(43)` == "Staying or living in a friend's room, apartment or house" |
             `Prior Living Situation(43)` == "Staying or living in a family member's room, apartment or house") &
             !is.na(ExitDestination)) %>%
    tabyl(ExitDestination) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "Entries", "Percent")),
  caption = "Count of Exits from Couching Entries"
  )

## show distribution of race for clients with any entry from incarceration during report period
knitr::kable(
  entries_during_period %>%
    filter(`Prior Living Situation(43)` == "Jail, prison or juvenile detention facility") %>%
    select(ClientID, RaceDesc) %>%
    distinct() %>%
    tabyl(RaceDesc, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of Race for Entries from Incarceration"
  )

knitr::kable(
  entries_during_period %>%
    filter(`Prior Living Situation(43)` == "Jail, prison or juvenile detention facility") %>%
    select(ClientID, HUDEthnicity) %>%
    distinct() %>%
    tabyl(HUDEthnicity, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of Ethnicity for Entries from Incarceration"
  )

```

% and number of people returning to homelessness (397 persons on 2019 SysPM)
- Demographics (gender, race, under 25 and 25 and over) and any known living situations, length of time at shelter or project before leaving? (is this possible within the system?)

Returns calculations do not include information for homelessness prevention, as this program type should not target literally homeless people. These calculations look at whether people exiting to a permanent destination between 10/1/17 and 9/30/18 returned to homelessness within two years of their first permanent exit in that period. They are higher than the SPM numbers (555 for this same period, according to the canned HMIS report run on 12.23.20) because they count re-entering coordinated entry as a return to homelessness.

Note: these percentages are by row.
```{r returns}

##  set program types again
return_program_types <- c("Transitional housing", "PH - Permanent Supportive Housing (disability required for entry)",
                          "PH – Housing Only", "PH – Housing with Services (no disability required for entry)",
                          "Emergency shelter", "Street outreach", "PH - Rapid Re-Housing", "Safe Haven",
                          "Coordinated Entry", "Diversion")

## get relevant exits with return flag
exited_two_years_ago <- all_data %>%
  left_join(return_flags, by = "EnrollID") %>%
  filter(ExitDate < ReportEnd - dyears(2) &
           ExitDate >= ReportStart - dyears(2) &
           ProgramType %in% return_program_types &
           !is.na(return_flag)) %>%
  group_by(ClientID) %>%
  arrange(ExitDate) %>%
  slice(1L) %>%
  ungroup() %>%
  mutate(return_flag = if_else(return_flag == 1, "Returned", "Did Not Return"))

##  show count of people by return status
knitr::kable(
  exited_two_years_ago %>%
    tabyl(return_flag, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People Returning"
  )

## show distribution of returns for clients by gender
knitr::kable(
  exited_two_years_ago %>%
    mutate(Gender = case_when(Gender == "Client refused" 
                     | Gender == "Data not collected" ~ "Unknown", 
                     Gender == "Trans Female (MTF or Male to Female)" ~ "Trans Female",
                     TRUE ~ Gender)) %>%
    tabyl(Gender, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Gender", ""),
  caption = "Count of People Returning by Gender"
  )

## show distribution of returns for clients by race and ethnicity
knitr::kable(
  exited_two_years_ago %>%
    tabyl(RaceDesc, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Race", ""),
  caption = "Count of People Returning by Race"
  )

knitr::kable(
  exited_two_years_ago %>%
    tabyl(HUDEthnicity, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Ethnicity", ""),
  caption = "Count of People Returning by Ethnicity"
  )

## show distribution of returns for clients by simple age group
knitr::kable(
  exited_two_years_ago %>%
    mutate(simple_age_group = if_else(`Age at Entry` < 25, "Under 25", "Over 25")) %>%
    tabyl(simple_age_group, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Age", ""),
  caption = "Count of People Returning by Age"
  )

## show distribution of returns for clients by original destination
knitr::kable(
  exited_two_years_ago %>%
    tabyl(ExitDestination, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Original Destination", ""),
  caption = "Count of People Returning by Destination"
  )

## show distribution of returns for clients by time in program
knitr::kable(
  exited_two_years_ago %>%
    mutate(time_in_program = as.duration(EnrollDate %--% ExitDate) / ddays(1),
           time_in_program = case_when(time_in_program < 15 ~ "14 days or less",
                                       time_in_program < 31 ~ "15-30 days",
                                       time_in_program < 91 ~ "31-90 days",
                                       time_in_program < 181 ~ "91-180 days",
                                       time_in_program < 365 ~ "181-365 days", 
                                       time_in_program < 731 ~ "366-730 days",
                                       TRUE ~ "731+ days"),
           time_in_program = factor(
             time_in_program,
             levels = c(
               "14 days or less", "15-30 days", "31-90 days", "91-180 days",
               "181-365 days",  "366-730 days", "731+ days")
    )) %>%
    tabyl(time_in_program, return_flag, show_na = FALSE) %>%
    adorn_totals("row") %>%
    adorn_percentages("row") %>%
    adorn_pct_formatting(0) %>%
    adorn_ns() %>%
    adorn_title("top", "Time in Program", ""),
  caption = "Count of People Returning by Time in Program"
  )

## get next "prior living situation" for each returning client
return_entry <- exited_two_years_ago %>%
  filter(return_flag == "Returned") %>%
  setNames(paste("Org", colnames(exited_two_years_ago), sep = "_")) %>%
  left_join(all_data, by = c("Org_ClientID" = "ClientID")) %>%
  filter(EnrollDate >= Org_ExitDate &
           Org_EnrollID != EnrollID) %>%
  arrange(EnrollDate) %>%
  group_by(Org_ClientID) %>%
  slice(1L) %>%
  ungroup()

## show distribution of returns for clients by prior residence at return
knitr::kable(
  return_entry %>%
    tabyl(`Prior Living Situation(43)`, show_na = FALSE) %>%
    adorn_pct_formatting(0) %>%
    setNames(c("", "People", "Percent")),
  caption = "Count of People Returning from Residence Type"
  )

```

